---
title: "Problem Set 6"
author: "Jayani Das 701"
date: "2026-03-13"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### To demonstrate the application of subset selection methods

Consider the “Hitters” data. For loading the data you need to install ISLR
library. We wish to predict a baseball player’s salary on the basis of various
predictors. So first we need to choose an optimal model.

```{r}
#install.packages("ISLR")
library(ISLR)
data(Hitters)
View(Hitters)
head(Hitters)
```
1. Clean the data by removing all rows that have missing value in any variable. You will be working with this clean data.
```{r}
hitters <- na.omit(Hitters)
dim(hitters)
```
2. Split the data randomly into train-test keeping 80% as training sample and remaining as test sample.
```{r}
set.seed(123)
train_index <- sample(1:nrow(hitters), size = 0.8 * nrow(hitters))
train_data <- hitters[train_index, ]; train_data
test_data <- hitters[-train_index, ]; test_data
```
3. Perform the best subset selection on the training sample. While choosing the best model fit up to a 19 variable model. Choose the best model according to adjusted R squared, BIC , AIC and Cp values. In case the best model differs according to the selection criteria, report the best fitted model according to each selection criterion.
```{r}
#install.packages("leaps")
library(leaps)
best_subset <- regsubsets(Salary ~ ., data = train_data, nvmax = 19)
summary_best_subset <- summary(best_subset); summary_best_subset

# Best model according to adjusted R squared
best_adj_r2 <- which.max(summary_best_subset$adjr2)
# Best model according to BIC
best_bic <- which.min(summary_best_subset$bic)
# Best model according to AIC
best_aic <- which.min(summary_best_subset$cp)
best_adj_r2
best_bic
best_aic

coefficients_adj_r2 <- coef(best_subset, best_adj_r2)
coefficients_bic <- coef(best_subset, best_bic)
coefficients_aic <- coef(best_subset, best_aic)
cat("Best model according to Adjusted R-squared includes variables:", names(coefficients_adj_r2), "\n")
cat("Best model according to BIC includes variables:", names(coefficients_bic), "\n")
cat("Best model according to AIC includes variables:", names(coefficients_aic), "\n")
```
##### REPORT:
- The best model according to Adjusted R-squared includes the following variables: (Intercept), AtBat, Hits, Walks, CAtBat, CHits, CWalks, LeagueN, DivisionW, NewLeagueN.

- The best model according to BIC includes the following variables: (Intercept), AtBat, Hits, Walks, CAtBat, CHits, CWalks, LeagueN, DivisionW, NewLeagueN.

- The best model according to AIC includes the following variables: (Intercept), AtBat, Hits, Walks, CAtBat, CHits, CWalks, LeagueN, DivisionW, NewLeagueN.


4. Find the training MSE and test MSE for each of the best models selected according to different criterion as mentioned in 3 and compare their values.
```{r}

# Function to calculate MSE for regsubsets models
calculate_mse <- function(model, data, id) {
  # Extract coefficients for the chosen subset
  coefi <- coef(model, id = id)
  
  # Build design matrix with the same variables
  mat <- model.matrix(Salary ~ ., data)[, names(coefi)]
  
  # Predictions
  predictions <- mat %*% coefi
  
  # MSE
  mse <- mean((data$Salary - predictions)^2)
  return(mse)
}

# Best model according to adjusted R squared
model_adj_r2 <- regsubsets(Salary ~ ., data = train_data, nvmax = best_adj_r2)
mse_train_adj_r2 <- calculate_mse(model_adj_r2, train_data, id = best_adj_r2)
mse_test_adj_r2  <- calculate_mse(model_adj_r2, test_data, id = best_adj_r2)

# Best model according to BIC
model_bic <- regsubsets(Salary ~ ., data = train_data, nvmax = best_bic)
mse_train_bic <- calculate_mse(model_bic, train_data, id = best_bic)
mse_test_bic  <- calculate_mse(model_bic, test_data, id = best_bic)


```
5. Repeat 3-4 using forward stepwise selection method.
```{r}
# Forward stepwise selection
forward_stepwise <- regsubsets(Salary ~ ., data = train_data, nvmax = 19, method = "forward")
summary_forward_stepwise <- summary(forward_stepwise); summary_forward_stepwise

# Best model according to adjusted R squared
best_adj_r2_forward <- which.max(summary_forward_stepwise$adjr2)

# Best model according to BIC
best_bic_forward <- which.min(summary_forward_stepwise$bic)

# Best model according to AIC
best_aic_forward <- which.min(summary_forward_stepwise$cp)

cat("Best model according to Adjusted R-squared (Forward):", names(coef(forward_stepwise, best_adj_r2_forward)), "\n")
cat("Best model according to BIC (Forward):", names(coef(forward_stepwise, best_bic_forward)), "\n")
cat("Best model according to AIC (Forward):", names(coef(forward_stepwise, best_aic_forward)), "\n")
```
6. Repeat 3-4 using backward stepwise selection method
```{r}
# Backward stepwise selection
backward_stepwise <- regsubsets(Salary ~ ., data = train_data, nvmax = 19, method = "backward")
summary_backward_stepwise <- summary(backward_stepwise)
summary_backward_stepwise

# Best model according to Adjusted R-squared
best_adj_r2_backward <- which.max(summary_backward_stepwise$adjr2)

# Best model according to BIC
best_bic_backward <- which.min(summary_backward_stepwise$bic)

# Best model according to AIC (using Mallows' Cp as proxy)
best_aic_backward <- which.min(summary_backward_stepwise$cp)

# Display selected variables for each criterion
cat("Best model according to Adjusted R-squared (Backward):",
    names(coef(backward_stepwise, id = best_adj_r2_backward)), "\n")

cat("Best model according to BIC (Backward):",
    names(coef(backward_stepwise, id = best_bic_backward)), "\n")

cat("Best model according to AIC (Backward):",
    names(coef(backward_stepwise, id = best_aic_backward)), "\n")


```
7. Compare the results of the three methods in terms of choice of best model, train MSE amd test MSE.
```{r}
# Function to calculate MSE for regsubsets models
calculate_mse <- function(model, data, id) {
  # Extract coefficients for the chosen subset
  coefi <- coef(model, id = id)
  
  # Build design matrix with the same variables
  mat <- model.matrix(Salary ~ ., data)[, names(coefi)]
  
  # Predictions
  predictions <- mat %*% coefi
  
  # MSE
  mse <- mean((data$Salary - predictions)^2)
  return(mse)
}

# Identify best models from forward stepwise summary
summary_forward_stepwise <- summary(forward_stepwise)
best_forward_adj_r2 <- which.max(summary_forward_stepwise$adjr2)
best_forward_bic    <- which.min(summary_forward_stepwise$bic)
best_forward_aic    <- which.min(summary_forward_stepwise$cp)  # Cp as proxy for AIC

# Compute MSEs
mse_train_forward_adj_r2 <- calculate_mse(forward_stepwise, train_data, id = best_forward_adj_r2)
mse_test_forward_adj_r2  <- calculate_mse(forward_stepwise, test_data, id = best_forward_adj_r2)

mse_train_forward_bic <- calculate_mse(forward_stepwise, train_data, id = best_forward_bic)
mse_test_forward_bic  <- calculate_mse(forward_stepwise, test_data, id = best_forward_bic)

mse_train_forward_aic <- calculate_mse(forward_stepwise, train_data, id = best_forward_aic)
mse_test_forward_aic  <- calculate_mse(forward_stepwise, test_data, id = best_forward_aic)

# Print results
cat("MSE for model selected by AIC (Forward): Train =", mse_train_forward_aic, "Test =", mse_test_forward_aic, "\n")
cat("MSE for model selected by Adjusted R-squared (Forward): Train =", mse_train_forward_adj_r2, "Test =", mse_test_forward_adj_r2, "\n")
cat("MSE for model selected by BIC (Forward): Train =", mse_train_forward_bic, "Test =", mse_test_forward_bic, "\n")


```

